E-Graphs With Bindings

### 2025-01-27 LICS_2025/e_graphs_with_bindings/sections/combinatorial_semantics.tex
clean-up

\subsection{DPOI-Rewriting for Hypergraphs with Interfaces}
DPO rewriting is a standard technique for formalising graph transformations categorically.
% In this section we recall standard results of double-pushout (DPO) rewriting of hypergraphs,  as well as the specific notion of DPO rewriting which is sound and complete for arbitrary SMTs,  which is called \textit{convex} rewriting.
The intuitive notion of (hyper-)graph rewriting is  as expected: given a rewrite rule $\mathcal L \rightsquigarrow \mathcal R$ and a graph $\mathcal{G}$,  we expect to find an occurrence of $\mathcal L$ within $\mathcal{G}$,  remove it,  and then plug $\mathcal R$ into the hole to recover the rewritten graph $\mathcal{H}$.  In DPO rewriting,  such a rewrite rule is formalised as a span $\mathcal L \xleftarrow{} \mathcal K \xrightarrow{} \mathcal R$ in some appropriate category of graphs,  where $\mathcal K$ is some invariant that should be maintained during rewriting.
A typical DPO square is depicted in Figure~\ref{fig:dpo_dpoi}~(left).

@@

To express the rewriting of (hyper-)graphs with interfaces,  \textit{i.e.},  the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout rewriting with \textit{interfaces}  (DPOI) as shown in Figure~\ref{fig:dpo_dpoi}~(right).
In the case of $ \catname{Hyp(\Sigma)}$,  we wish for a rewrite rules to be pairs of discrete cospans of hypergraphs with matching interfaces: $n \xrightarrow{} \mathcal L \xleftarrow{} m$, $n \xrightarrow{} \mathcal R \xleftarrow{} m$.  
By noting that a cospan $0 \xrightarrow{} \mathcal L \xleftarrow{[f,g]} n + m$ encodes the same data as a cospan $n \xrightarrow{f} \mathcal L \xleftarrow{g} m$ (Remark 3.14~\cite{bonchi_string_2022-1}) we can formulate a rewrite rule as a single span $\mathcal L \xleftarrow{} n+m \xrightarrow{} \mathcal R$,  and we can do similarly for the discrete cospans giving the interface to $\mathcal G$ and $\mathcal H$.  
We now make this identification throughout without further mention.  
Thus we reach the following variant of DPO rewriting.
\begin{definition}[DPOI rewriting]\label{def:dpoi}
Given a span of morphisms $\mathcal G \xleftarrow{} n+m \xrightarrow{} \mathcal H$ in $\textbf{Hyp}(\Sigma)$,  we say \textit{$\mathcal G$ rewrites to $\mathcal H$ with interface $n+m$ via rewrite rule $\mathcal L \xleftarrow{} i+j \xrightarrow{} \mathcal R$} if there exists an object $\mathcal C$ and morphisms which complete the rightmost commutative diagram in Figure~\ref{fig:dpo_dpoi} such that the two marked squares are pushouts.

@@

\begin{definition}
We define an e-hypergraph $\mathcal{G}$ over a closed symmetric monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$ as a tuple $(V,E,s,t, l_{V}, l_{E} \textcolor{orange}{<}, \textcolor{blue}{<},\consistency)$ where
\begin{itemize}
  \item $V$ is a set of vertices.
  \item $E = \colorbox{yellow}{E} \cup \textcolor{orange}{E} \cup \textcolor{blue}{E}$ is a set of hyperedges that that is formed of three disjoint sets of hyperedges defined below.
  \item $s,t : E \to V^{*}$ are source and target functions.
  \item $l_{E} : E \to \Sigma_{M} + 1$, $l_{V} : V \to \Sigma_{O} \cup \text{obj}_{\multimap}$ are \textit{label} functions, where $\text{obj}_{\multimap}$ consists of objects $A \multimap B$ for $A,B \in obj_{\Sigma_{O}}$.
  \item $\textcolor{orange}{<} \subseteq \textcolor{orange}{E} \times V + E$ and $\textcolor{blue}{<} \subseteq \textcolor{blue}{E} \times V + E $ with $\textcolor{orange}{E}, \textcolor{blue} \subset E$ are disjoint partial orders that are defined as a transitive closure of corresponding immediate predecessor relations that we denote as $\textcolor{orange}{<^{\mu}}$ and $\textcolor{blue}{<^{\mu}}$.
        For $x \in V + E$, the set of predecessors of $x$ is denoted as $[x) = \{x' ~|~ \exists y_{1} = x', \ldots, y_{n} = x,  y_{i} \textcolor{orange}{<^{\mu}} y_{i + 1} \text{ or } y_{i} \textcolor{blue}{<^{\mu}} y_{i+1}\; \}$.
        We call edges $e$ such that $l(e) = \bot$ hierarchical and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \textit{top-level}.
        Each of these relations should satisfy the following
        \begin{enumerate}
          \item each set of predecessors consists exclusively of hierarchical edges;
          \item each $x$ has at most one immediate predecessor;
          \item edges $e$ such that there is no $x$ such that $e \textcolor{orange}{<^{\mu}} x$ or $e \textcolor{orange}{<^{\mu}} x$ are labelled;
          \item the relations are closed under connectivity, \emph{i.e.}, if $v \in s(e)$ then $e' \textcolor{orange}{<^{\mu}} e$ iff $e' \textcolor{orange}{<^{\mu}} v$, similarly for $v \in t(e)$ and $\textcolor{blue}{<}$.
        \end{enumerate}
  \item $\consistency$ is a \textit{consistency relation} which is given by the union of a family of equivalence relations $\consistency_p$ on each set $\{x \in V + E ~|~ p \textcolor{orange}{<^\mu} x\}$ of elements which share the same parent where each relation is also closed under connectivity, \textit{i.e.}, if $v \in s(e)$ or $v \in t(e)$ such that $p \textcolor{orange}{<^{\mu}}(v)$ and $p \textcolor{orange}{<^{\mu}}(e)$ then $v \consistency_{p} e$.
  We require that $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$ where $V_{p} = \{ v ~ | ~ p \textcolor{orange}{<^{\mu}} v\}$ and $E_{p} = \{ e ~ | ~ p \textcolor{orange}{<^{\mu}} e\}$.
\end{itemize} 
$\textcolor{orange}{E}$ and $\textcolor{blue}{E}$ are then induced by the corresponding relations and consist of exclusively hierarchical edges, so $\colorbox{yellow}{E} = E \setminus \textcolor{orange}{E} \cup \textcolor{blue}{E}$.
Given an e-hypergraph $\mathcal{G}$ we can consider a corresponding \textit{underlying} hypergraph $\mathcal{G}'$ by forgetting $\textcolor{orange}{<}, \textcolor{blue}{<}$ and~$\consistency$.
\end{definition}

Intuitively, edges from $\textcolor{orange}{E}$ encode equivalence classes, \emph{i.e.}, e-boxes, while edges from $\textcolor{blue}{E}$ encode lambda-abstraction boxes.
Also note the typing on the $l_{V}$ function.
We work with freely generated closed semilattice-enriched categories, and to avoid unnecessary complications we limit the labelling of vertices to such a type.
In particular, it prohibits vertices to be labelled with, \emph{e.g.}, $A \otimes B$, but allows them to be labelled with $A \otimes B \multimap C \otimes D$.
$\textcolor{blue}{<}$ defines closed monoidal structure as defined in~\cite{fscd} and $(\textcolor{orange}{<},\consistency)$ defines $\catname{SLat}$ structure as per~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.

Consider an example in the middle part of Figure~\ref{fig:e-cospan-example}.
The e-hypergraph in the example is given by $V = \{v_{1}, \ldots w_{10}\}$, $E = \{e_{1}, \ldots, e_{7}\}$, $\textcolor{orange}{E} = \{e_{1}\}$, $\textcolor{blue}{E} = \{e_{2}\}$, $l_{E} = {e_{1} \mapsto \bot, e_{2} \mapsto \bot, e_{3} \mapsto g, \ldots, e_{7} \mapsto f}$ and then we have
\[
\vspace{-1ex}
\begin{array}{cccc}
  v_{4} \textcolor{orange}{<^{\mu}} e_{1} & v_{7} \textcolor{blue}{<^{\mu}} e_{2} & v_{4} \consistency e_{2} & v_{4} \not \consistency v_{10}\\
  e_{3} \textcolor{orange}{<^{\mu}} e_{1} & v_{8} \textcolor{blue}{<^{\mu}} e_{2} &  e_{2} \consistency u_{1} & e_{2} \not \consistency e_{6}\\
  e_{2} \textcolor{orange}{<^{\mu}} e_{1} & e_{5} \textcolor{blue}{<^{\mu}} e_{2} & e_{2} \consistency e_{4} & e_{4} \not \consistency e_{7}\\  
  \ldots & \ldots & \ldots & \ldots \\
  v_{12} \textcolor{orange}{<^{\mu}} e_{1} & w_{9} \textcolor{blue}{<^{\mu}} e_{2} & v_{11} \consistency v_{12} & w_{6} \not \consistency w_{10}
\end{array}
\]
and, in particular, we have $[e_{5}) = \{e_2, e_1\}$.
$\consistency_{e_{1}}$ defines a partition on immediate successors of $e_{1}$ which we delimit by a vertical dashed line.
Edges in $\textcolor{orange}{E}$ are depicted with a dashed box, while edges in $\textcolor{blue}{E}$ --- with a round solid box.
\begin{figure}
\[
\adjustbox{scale=0.6}{
\tikzfig{./figures/closed_iso_example_2}
}
\]
\captionsetup{skip=0pt, belowskip=-1ex}
\caption{Cospan of e-hypergraphs example}
\label{fig:e-cospan-example}
\end{figure}
\begin{definition}[E-hypergraph homomorphism]
  An \textit{e-hypergraph homomorphism} $\phi:\mathcal{F}\to\mathcal{G}$ between e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that
  \begin{enumerate}
      \item $\phi$ is a hypergraph homomorphism;
      \item $\phi_{E}(\textcolor{orange}{E_{\mathcal{F}}}) \subseteq \textcolor{orange}{E_{\mathcal{G}}}$ and $\phi_{E}(\textcolor{blue}{E_{\mathcal{F}}}) \subseteq \textcolor{blue}{E_{\mathcal{G}}}$;
      \item for $v \in V_{\mathcal{F}}$ and $e \in E_{\mathcal{F}}$
        \begin{align*}
          \text{if } e \textcolor{orange}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{orange}{<_{\mathcal{G}}^{\mu}} \phi(v)
          \qquad
          \text{if } e \textcolor{blue}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{blue}{<_{\mathcal{G}}^{\mu}} \phi(v)
        \end{align*}
          and for $e_1, e_2 \in E_{\mathcal{F}}$
          \begin{align*}
          \text{if } e_1 \textcolor{orange}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{orange}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \qquad
          \text{if } e_1 \textcolor{blue}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{blue}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \end{align*}
          \item for all $x_1, x_2 \in \mathcal{F}$,  $x_1 ~\consistency_{\mathcal{F}}~ x_2$ implies $\phi(x_1) ~\consistency_{\mathcal{G}}~ \phi(x_2)$.
  \end{enumerate}
  % If we further have that if $[x_1) = \varnothing$ then $[\phi(x_1)) = \varnothing$ we call the homomorphism \textit{strict}.
  \end{definition}
  The conditions on e-hypergraph homomorphisms require preserving immediate predecessors and the consistency relation.
  \begin{definition}[Category of e-hypergraphs]
  The \emph{category of e-hypergraphs},  denoted  $\catname{EHyp(\Sigma)}$,  has e-hypergraphs as objects and e-hypergraph homomorphisms as morphisms.  
  \end{definition}

@@

\end{definition}
We call $n$  \textit{external input interfaces}, $n'$ \textit{internal input interfaces},
and $n' \setminus f_{ext}(n)$  the \textit{strictly internal input interfaces}.  
We do analogously for the \emph{output interfaces},  with respect to $m$,  $m'$ and $m' \setminus g_{ext}(m)$.  
We occasionally conflate $f_{ext}$ with $f_{ext};f_{int}$ when it is clear from context,  and also conflate $n$ and $m$ with their images in $n'$ and $m'$,  and likewise $m, m, n', m'$ with their images in $\mathcal{G}$. 
Given an edge $e \in E_\mathcal{G}$ such that $l(e) = \bot$,  we call the \textit{inputs of $e$} the intersection of the strictly internal input interface of $\mathcal{G}$ with the immediate successors of $e$,  and analogously for the \emph{outputs of $e$}.
Graphically, we depict such extended cospans as in Figure~\ref{fig:e-cospan-example}: feet of the cospan are depicted within shaded regions.
All non-black vertices form strictly internal interfaces.

\begin{definition}[Isomorphic cospans]
Consider a relation 
\begin{multline*}
R = \bigl\{ x R y \text{ if } f_{\text{int}}(x) \consistency f_{\text{int}}(y)\\ \;\text{or}\; \exists z \;\text{s.t.}\; z \textcolor{blue}{<^{\mu}} f_{\text{int}}(x) \text{ and } z \textcolor{blue}{<}^{\mu} f_{\text{int}}(y)\bigr\}
\end{multline*}
for $x, y \in n'$.
And let $S$ be its reflexive closure.
The latter partitions $n'$ into non-empty subsets $\{p_{j}\}^{k}_{j=1}$.
We get an analogous partition for $m'$.

@@

In Figure~\ref{fig:e-cospan-example} the partitions for input and output internal interfaces are given with colours.

Intuitively, such notion of isomorphism makes it possible to freely rearrange interfaces for disjoint components.
For example, in the input interface of cospan in Figure~\ref{fig:e-cospan-example}, we can swap 
\begin{tikzpicture}
\begin{pgfonlayer}{nodelayer}
		\node [style=red node, label={above:$v_{4}$}] (0) at (-0.5, 3.5) {};
		\node [style=red node, label={above:$v_5$}] (1) at (0, 3.5) {};
		\node [style=red node, label={above:$v_6$}] (2) at (0.5, 3.5) {};
	\end{pgfonlayer}
\end{tikzpicture}
with 
\begin{tikzpicture}
\begin{pgfonlayer}{nodelayer}
    \node [style=blue node, label={above:$v_{7}$}] (0) at (-0.5, 3.5) {};
    \node [style=blue node, label={above:$v_8$}] (1) at (0, 3.5) {};
    \node [style=blue node, label={above:$v_9$}] (2) at (0.5, 3.5) {};
\end{pgfonlayer}
\end{tikzpicture}
but we can not swap
\begin{tikzpicture}
    \begin{pgfonlayer}{nodelayer}
            \node [style=red node, label={above:$v_{4}$}] (0) at (-0.5, 3.5) {};
            \node [style=red node, label={above:$v_5$}] (1) at (0, 3.5) {};
        \end{pgfonlayer}
\end{tikzpicture}
with each other as it would result in a non-isomorphic cospan.

@@

\begin{definition}
  We call a monogamous cospan
  \[
    n \xrightarrow{f_{\text{ext}}} n' \xrightarrow{f_{\text{int}}} \mathcal{G} \xleftarrow{g'_{\text{int}}} m' \xleftarrow{g_{\text{ext}}} m
  \]
  \textit{well-typed} if all hierarchical edges of $\mathcal{G}$ are well-typed in the sense below.
  \begin{itemize}
    \item For each $e \in \textcolor{orange}{E}$ consider sets $I$ and $O$ of its input and output vertices partitioned according to $\consistency_{e}$.
          $e$ is well-typed if for each element $S$ of the partition of $I$ $w(S) = w(s(e))$ and similarly for $O$ and $t(e)$.
    \item For each $e \in \textcolor{blue}{E}$ consider sets $I$ and $O$ of its input and output vertices.
          $e$ is well-typed if there exists an object $B \in \text{obj}_{\Sigma_{O}}$ such that $[w(s(e)), B] = w(I)$ and $B \multimap w(O) = w(t(e))$.
  \end{itemize}

@@

Consider, again, Figure ~\ref{fig:e-cospan-example}.
To make edge $e_{1}$ well-typed, it is necessary that $l_{V}^{*}([v_{4},v_{5},v_{6}]) = l_{V}^{*}([v_{10},v_{11},v_{12}]) = l_{V}^{*}([v_1,v_2,v_3])$ and $l_{V}^{*}([w_{4},w_{5},w_{6}]) = l_{V}^{*}([w_{10},v_{12},v_{11}]) = l_{V}^{*}([w_1,w_2,w_3])$ (note the change of ordering in the output of $e_{1}$).
To make edge $e_{2}$ well-typed, it is necessary that $l_{V}^{*}([v_4,v_5,v_9]) = l_{V}^{*}([v_7,v_8,v_9])$ and $l_{V}([u_1]) = l_{V}([v_9]) \multimap l_{V}^{*}([v_8,v_7,w_9])$.

@@

% Another concern is when we delete the occurrence of the left-hand side of a rewrite rule from an e-hypergraph we also modify this e-hypergraph's interfaces.
% Consider a hypothetical DPO square below in Figure~\ref{fig:interface_change_example} where long arrows denote matching between carriers of cospans.

>>>>

\subsection{DPOI-Rewriting for Hypergraphs with Interfaces}
DPO rewriting is a standard technique for formalising graph transformations categorically.
% In this section we recall standard results of double-pushout (DPO) rewriting of hypergraphs,  as well as the specific notion of DPO rewriting which is sound and complete for arbitrary SMTs,  which is called \textit{convex} rewriting.
The intuitive notion of (hyper-)graph rewriting is  as expected: given a rewrite rule $\mathcal L \rightsquigarrow \mathcal R$ and a graph $\mathcal{G}$,  we expect to find an occurrence of $\mathcal L$ within $\mathcal{G}$,  remove it,  and then plug $\mathcal R$ into the hole to recover the rewritten graph $\mathcal{H}$.  In DPO rewriting,  such a rewrite rule is formalised as a span $\mathcal L \xleftarrow{} \mathcal K \xrightarrow{} \mathcal R$ in some appropriate category of graphs,  where $\mathcal K$ is some invariant that should be maintained during rewriting.
A typical DPO square is depicted in Fig.~\ref{fig:dpo_dpoi}~(left).

@@

To express the rewriting of (hyper-)graphs with interfaces,  \textit{i.e.},  the ones that represent morphisms in $\HypI{\Sigma}$, the DPO formalism has been extended to double-pushout rewriting with \textit{interfaces}  (DPOI) as shown in Fig.~\ref{fig:dpo_dpoi}~(right).
In the case of $ \catname{Hyp(\Sigma)}$,  we wish for a rewrite rules to be pairs of discrete cospans of hypergraphs with matching interfaces: $n \xrightarrow{} \mathcal L \xleftarrow{} m$, $n \xrightarrow{} \mathcal R \xleftarrow{} m$.  
By noting that a cospan $0 \xrightarrow{} \mathcal L \xleftarrow{[f,g]} n + m$ encodes the same data as a cospan $n \xrightarrow{f} \mathcal L \xleftarrow{g} m$ (Remark 3.14~\cite{bonchi_string_2022-1}) we can formulate a rewrite rule as a single span $\mathcal L \xleftarrow{} n+m \xrightarrow{} \mathcal R$,  and we can do similarly for the discrete cospans giving the interface to $\mathcal G$ and $\mathcal H$.  
We now make this identification throughout without further mention.  
Thus we reach the following variant of DPO rewriting.
\begin{definition}[DPOI rewriting]\label{def:dpoi}
Given a span of morphisms $\mathcal G \xleftarrow{} n+m \xrightarrow{} \mathcal H$ in $\textbf{Hyp}(\Sigma)$,  we say \textit{$\mathcal G$ rewrites to $\mathcal H$ with interface $n+m$ via rewrite rule $\mathcal L \xleftarrow{} i+j \xrightarrow{} \mathcal R$} if there exists an object $\mathcal C$ and morphisms which complete the rightmost commutative diagram in Fig.~\ref{fig:dpo_dpoi} such that the two marked squares are pushouts.

@@

\begin{definition}
We define an e-hypergraph $\mathcal{G}$ over a closed symmetric monoidal signature $\Sigma = (\Sigma_{O}, \Sigma_{M})$ as a tuple $(V,E,s,t, l_{V}, l_{E} \textcolor{e-color}{<}, \textcolor{closed-color}{<},\consistency)$ where
\begin{itemize}
  \item $V$ is a set of vertices.
  \item $E = \colorbox{yellow}{E} \cup \textcolor{e-color}{E} \cup \textcolor{closed-color}{E}$ is a set of hyperedges that that is formed of three disjoint sets of hyperedges defined below.
  \item $s,t : E \to V^{*}$ are source and target functions.
  \item $l_{E} : E \to \Sigma_{M} + 1$, $l_{V} : V \to \Sigma_{O} \cup \text{obj}_{\multimap}$ are \textit{label} functions, where $\text{obj}_{\multimap}$ consists of objects $A \multimap B$ for $A,B \in obj_{\Sigma_{O}}$.
  \item $\textcolor{e-color}{<} \subseteq \textcolor{e-color}{E} \times V + E$ and $\textcolor{closed-color}{<} \subseteq \textcolor{closed-color}{E} \times V + E $ with $\textcolor{e-color}{E}, \textcolor{closed-color} \subset E$ are disjoint partial orders that are defined as a transitive closure of corresponding immediate predecessor relations that we denote as $\textcolor{e-color}{<^{\mu}}$ and $\textcolor{closed-color}{<^{\mu}}$.
        For $x \in V + E$, the set of predecessors of $x$ is denoted as $[x) = \{x' ~|~ \exists y_{1} = x', \ldots, y_{n} = x,  y_{i} \textcolor{e-color}{<^{\mu}} y_{i + 1} \text{ or } y_{i} \textcolor{closed-color}{<^{\mu}} y_{i+1}\; \}$.
        We call edges $e$ such that $l(e) = \bot$ hierarchical and edges $e$ and vertices $v$ such that $[e) = \varnothing$ and $[v) = \varnothing$ \textit{top-level}.
        Each of these relations should satisfy the following
        \begin{enumerate}
          \item each set of predecessors consists exclusively of hierarchical edges;
          \item each $x$ has at most one immediate predecessor;
          \item edges $e$ such that there is no $x$ such that $e \textcolor{e-color}{<^{\mu}} x$ or $e \textcolor{e-color}{<^{\mu}} x$ are labelled;
          \item the relations are closed under connectivity, \emph{i.e.}, if $v \in s(e)$ then $e' \textcolor{e-color}{<^{\mu}} e$ iff $e' \textcolor{e-color}{<^{\mu}} v$, similarly for $v \in t(e)$ and $\textcolor{closed-color}{<}$.
        \end{enumerate}
  \item $\consistency$ is a \textit{consistency relation} which is given by the union of a family of equivalence relations $\consistency_p$ on each set $\{x \in V + E ~|~ p \textcolor{e-color}{<^\mu} x\}$ of elements which share the same parent where each relation is also closed under connectivity, \textit{i.e.}, if $v \in s(e)$ or $v \in t(e)$ such that $p \textcolor{e-color}{<^{\mu}}(v)$ and $p \textcolor{e-color}{<^{\mu}}(e)$ then $v \consistency_{p} e$.
  We require that $\consistency_{p} \not = (E_{p} + V_{p}) \times (E_{p} + V_{p})$ where $V_{p} = \{ v ~ | ~ p \textcolor{e-color}{<^{\mu}} v\}$ and $E_{p} = \{ e ~ | ~ p \textcolor{e-color}{<^{\mu}} e\}$.
\end{itemize} 
$\textcolor{e-color}{E}$ and $\textcolor{closed-color}{E}$ are then induced by the corresponding relations and consist of exclusively hierarchical edges, so $\colorbox{yellow}{E} = E \setminus \textcolor{e-color}{E} \cup \textcolor{closed-color}{E}$.
Given an e-hypergraph $\mathcal{G}$ we can consider a corresponding \textit{underlying} hypergraph $\mathcal{G}'$ by forgetting $\textcolor{e-color}{<}, \textcolor{closed-color}{<}$ and~$\consistency$.
\end{definition}

Intuitively, edges from $\textcolor{e-color}{E}$ encode equivalence classes, \emph{i.e.}, e-boxes, while edges from $\textcolor{closed-color}{E}$ encode lambda-abstraction boxes.
Also note the typing on the $l_{V}$ function.
We work with freely generated closed semilattice-enriched categories, and to avoid unnecessary complications we limit the labelling of vertices to such a type.
In particular, it prohibits vertices to be labelled with, \emph{e.g.}, $A \otimes B$, but allows them to be labelled with $A \otimes B \multimap C \otimes D$.
$\textcolor{closed-color}{<}$ defines closed monoidal structure as defined in~\cite{fscd} and $(\textcolor{e-color}{<},\consistency)$ defines $\catname{SLat}$ structure as per~\cite{ghica2024equivalencehypergraphsegraphsmonoidal}.

Consider an example in the middle part of Fig.~\ref{fig:e-cospan-example}.
The e-hypergraph in the example is given by $V = \{v_{1}, \ldots w_{10}\}$, $E = \{e_{1}, \ldots, e_{7}\}$, $\textcolor{e-color}{E} = \{e_{1}\}$, $\textcolor{closed-color}{E} = \{e_{2}\}$, $l_{E} = {e_{1} \mapsto \bot, e_{2} \mapsto \bot, e_{3} \mapsto g, \ldots, e_{7} \mapsto f}$ and then we have
\[
\vspace{-1ex}
\begin{array}{cccc}
  v_{4} \textcolor{e-color}{<^{\mu}} e_{1} & v_{7} \textcolor{closed-color}{<^{\mu}} e_{2} & v_{4} \consistency e_{2} & v_{4} \not \consistency v_{10}\\
  e_{3} \textcolor{e-color}{<^{\mu}} e_{1} & v_{8} \textcolor{closed-color}{<^{\mu}} e_{2} &  e_{2} \consistency u_{1} & e_{2} \not \consistency e_{6}\\
  e_{2} \textcolor{e-color}{<^{\mu}} e_{1} & e_{5} \textcolor{closed-color}{<^{\mu}} e_{2} & e_{2} \consistency e_{4} & e_{4} \not \consistency e_{7}\\  
  \ldots & \ldots & \ldots & \ldots \\
  v_{12} \textcolor{e-color}{<^{\mu}} e_{1} & w_{9} \textcolor{closed-color}{<^{\mu}} e_{2} & v_{11} \consistency v_{12} & w_{6} \not \consistency w_{10}
\end{array}
\]
and, in particular, we have $[e_{5}) = \{e_2, e_1\}$.
$\consistency_{e_{1}}$ defines a partition on immediate successors of $e_{1}$ which we delimit by a vertical dashed line.
Edges in $\textcolor{e-color}{E}$ are depicted with a dashed box, while edges in $\textcolor{closed-color}{E}$ --- with a round solid box.
\begin{figure}
\[
\adjustbox{scale=0.6}{
\tikzfig{./figures/closed_iso_example_2}
}
\]
\captionsetup{skip=0pt, belowskip=-1ex}
\caption{Cospan of e-hypergraphs example}
\label{fig:e-cospan-example}
\end{figure}
\begin{definition}[E-hypergraph homomorphism]
  An \textit{e-hypergraph homomorphism} $\phi:\mathcal{F}\to\mathcal{G}$ between e-hypergraphs $\mathcal{F},\mathcal{G}$ is a pair of functions $\phi_V : V_{\mathcal{F}} \to V_{\mathcal{G}}, \phi_E : E_{\mathcal{F}} \to E_{\mathcal{G}}$ such that
  \begin{enumerate}
      \item $\phi$ is a hypergraph homomorphism;
      \item $\phi_{E}(\textcolor{e-color}{E_{\mathcal{F}}}) \subseteq \textcolor{e-color}{E_{\mathcal{G}}}$ and $\phi_{E}(\textcolor{closed-color}{E_{\mathcal{F}}}) \subseteq \textcolor{closed-color}{E_{\mathcal{G}}}$;
      \item for $v \in V_{\mathcal{F}}$ and $e \in E_{\mathcal{F}}$
        \begin{align*}
          \text{if } e \textcolor{e-color}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{e-color}{<_{\mathcal{G}}^{\mu}} \phi(v)
          \qquad
          \text{if } e \textcolor{closed-color}{<_{\mathcal{F}}^{\mu}} v \text{ then } \phi(e) \textcolor{closed-color}{<_{\mathcal{G}}^{\mu}} \phi(v)
        \end{align*}
          and for $e_1, e_2 \in E_{\mathcal{F}}$
          \begin{align*}
          \text{if } e_1 \textcolor{e-color}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{e-color}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \qquad
          \text{if } e_1 \textcolor{closed-color}{<_{\mathcal{F}}^{\mu}} e_2 \text{ then } \phi(e_1) \textcolor{closed-color}{<_{\mathcal{G}}^{\mu}} \phi(e_2)
          \end{align*}
          \item for all $x_1, x_2 \in \mathcal{F}$,  $x_1 ~\consistency_{\mathcal{F}}~ x_2$ implies $\phi(x_1) ~\consistency_{\mathcal{G}}~ \phi(x_2)$.
  \end{enumerate}
  % If we further have that if $[x_1) = \varnothing$ then $[\phi(x_1)) = \varnothing$ we call the homomorphism \textit{strict}.
  \end{definition}
  The conditions on e-hypergraph homomorphisms require preserving immediate predecessors and the consistency relation.
  \begin{definition}[Category of e-hypergraphs]
  The \emph{category of e-hypergraphs},  denoted  $\catname{EHyp(\Sigma)}$,  has e-hypergraphs as objects and e-hypergraph homomorphisms as morphisms.  
  \end{definition}

@@

\end{definition}
We call $n$  \textit{external input interfaces}, $n'$ \textit{internal input interfaces},
and $n' \setminus f_{ext}(n)$  the \textit{strictly internal input interfaces}.  
We do analogously for the \emph{output interfaces},  with respect to $m$,  $m'$ and $m' \setminus g_{ext}(m)$.  
We occasionally conflate $f_{ext}$ with $f_{ext};f_{int}$ when it is clear from context,  and also conflate $n$ and $m$ with their images in $n'$ and $m'$,  and likewise $m, m, n', m'$ with their images in $\mathcal{G}$. 
Given an edge $e \in E_\mathcal{G}$ such that $l(e) = \bot$,  we call the \textit{inputs of $e$} the intersection of the strictly internal input interface of $\mathcal{G}$ with the immediate successors of $e$,  and analogously for the \emph{outputs of $e$}.
Graphically, we depict such extended cospans as in Fig.~\ref{fig:e-cospan-example}: feet of the cospan are depicted within shaded regions.
All non-black vertices form strictly internal interfaces.

\begin{definition}[Isomorphic cospans]
Consider a relation 
\begin{multline*}
R = \bigl\{ x R y \text{ if } f_{\text{int}}(x) \consistency f_{\text{int}}(y)\\ \;\text{or}\; \exists z \;\text{s.t.}\; z \textcolor{closed-color}{<^{\mu}} f_{\text{int}}(x) \text{ and } z \textcolor{closed-color}{<}^{\mu} f_{\text{int}}(y)\bigr\}
\end{multline*}
for $x, y \in n'$.
And let $S$ be its reflexive closure.
The latter partitions $n'$ into non-empty subsets $\{p_{j}\}^{k}_{j=1}$.
We get an analogous partition for $m'$.

@@

In Fig.~\ref{fig:e-cospan-example} the partitions for input and output internal interfaces are given with colours.

Intuitively, such notion of isomorphism makes it possible to freely rearrange interfaces for disjoint components.
For example, in the input interface of cospan in Fig.~\ref{fig:e-cospan-example}, we can swap 
\begin{tikzpicture}
\begin{pgfonlayer}{nodelayer}
		\node [style=red node, label={above:$v_{4}$}] (0) at (-0.5, 3.5) {};
		\node [style=red node, label={above:$v_5$}] (1) at (0, 3.5) {};
		\node [style=red node, label={above:$v_6$}] (2) at (0.5, 3.5) {};
	\end{pgfonlayer}
\end{tikzpicture}
with 
\begin{tikzpicture}
\begin{pgfonlayer}{nodelayer}
    \node [style=blue node, label={above:$v_{7}$}] (0) at (-0.5, 3.5) {};
    \node [style=blue node, label={above:$v_8$}] (1) at (0, 3.5) {};
    \node [style=blue node, label={above:$v_9$}] (2) at (0.5, 3.5) {};
\end{pgfonlayer}
\end{tikzpicture}
but we can not swap
\begin{tikzpicture}
    \begin{pgfonlayer}{nodelayer}
            \node [style=red node, label={above:$v_{4}$}] (0) at (-0.5, 3.5) {};
            \node [style=red node, label={above:$v_5$}] (1) at (0, 3.5) {};
        \end{pgfonlayer}
\end{tikzpicture}
with each other as it would result in a non-isomorphic cospan.

@@

\begin{definition}
  We call a monogamous cospan
  \[
    n \xrightarrow{f_{\text{ext}}} n' \xrightarrow{f_{\text{int}}} \mathcal{G} \xleftarrow{g'_{\text{int}}} m' \xleftarrow{g_{\text{ext}}} m
  \]
  \textit{well-typed} if all hierarchical edges of $\mathcal{G}$ are well-typed in the sense below.
  \begin{itemize}
    \item For each $e \in \textcolor{e-color}{E}$ consider sets $I$ and $O$ of its input and output vertices partitioned according to $\consistency_{e}$.
          $e$ is well-typed if for each element $S$ of the partition of $I$ $w(S) = w(s(e))$ and similarly for $O$ and $t(e)$.
    \item For each $e \in \textcolor{closed-color}{E}$ consider sets $I$ and $O$ of its input and output vertices.
          $e$ is well-typed if there exists an object $B \in \text{obj}_{\Sigma_{O}}$ such that $[w(s(e)), B] = w(I)$ and $B \multimap w(O) = w(t(e))$.
  \end{itemize}

@@

Consider, again, Fig. ~\ref{fig:e-cospan-example}.
To make edge $e_{1}$ well-typed, it is necessary that $l_{V}^{*}([v_{4},v_{5},v_{6}]) = l_{V}^{*}([v_{10},v_{11},v_{12}]) = l_{V}^{*}([v_1,v_2,v_3])$ and $l_{V}^{*}([w_{4},w_{5},w_{6}]) = l_{V}^{*}([w_{10},v_{12},v_{11}]) = l_{V}^{*}([w_1,w_2,w_3])$ (note the change of ordering in the output of $e_{1}$).
To make edge $e_{2}$ well-typed, it is necessary that $l_{V}^{*}([v_4,v_5,v_9]) = l_{V}^{*}([v_7,v_8,v_9])$ and $l_{V}([u_1]) = l_{V}([v_9]) \multimap l_{V}^{*}([v_8,v_7,w_9])$.

@@

% Another concern is when we delete the occurrence of the left-hand side of a rewrite rule from an e-hypergraph we also modify this e-hypergraph's interfaces.
% Consider a hypothetical DPO square below in Fig.~\ref{fig:interface_change_example} where long arrows denote matching between carriers of cospans.
